---
title: 'Reproducible Research: Peer Assessment 1'
author: "sirir"
date: "Saturday, November 15, 2014"
output: html_document
---
####Loading and preprocessing the data
Loading data
```{r, echo = TRUE}
data <- read.csv("./activity.csv",sep = ",",header = TRUE, na.strings ="NA",
                   colClasses = c('integer','Date','factor'))
```
Summary of Original data
```{r, echo=TRUE}
summary(data)
```
First five lines of Original data
```{r, echo=TRUE}
head(data)
```
Loading Libraries

```{r, echo=FALSE}
library(ggplot2)
```

Removing missing values from the original data and assigning it to 'processedData'
variable
```{r, echo=TRUE}
processedData <- na.omit(data)
```
First five lines of Processed Data
```{r, echo=TRUE}
head(processedData)
```
Summary of processed Data
```{r, echo=TRUE}
summary(processedData)
```
Total number of steps per day and first five days
```{r}
steps.per.day <- tapply(processedData$steps, processedData$date, FUN = sum)
head(steps.per.day)
```
Histogram of total number of steps per day.

```{r}
plot1 <- ggplot(processedData, aes(date, steps)) + geom_bar(stat = "identity",
        binwidth = .5) + labs(title = "Histogram of Total Number of Steps taken per Day", x = "Date", y = "Total Number of Steps")
print(plot1)
```

Mean total number of steps taken per day
```{r}
mean(total.steps)
```
Median total number of steps taken per day
```{r}
median(total.steps)
```
Average daily activity pattern
```{r}
averages <- aggregate(processedData$steps, list(interval = as.numeric(as.character(processedData$interval))), FUN = "mean")
names(averages)[2] <- "Avg.Steps"

plot2 <- ggplot(averages, aes(interval, Avg.Steps)) + geom_line(color = "blue", size = 0.7) + labs(title = "Time Series Plot of the 5-minute Intervals", x = "5-minute intervals", y = "Average Number of Steps Taken")
print(plot2)
```

The 5-minute inteval with maximum steps across all the day
```{r}
averages[averages$Avg.Steps == max(averages$Avg.Steps),]
```

####Imputing missing values
The total number of missing values in the dataset
```{r}
sum(!complete.cases(data))
```
Imputing missing values by using mean for that 5-minute interval
```{r}
impData <- data 
for (i in 1:nrow(impData)) {
    if (is.na(impData$steps[i])) {
        impData$steps[i] <- averages[which(impData$interval[i] == averages$interval), ]$Avg.Steps
    }
}
```

Number of missing values after after imputation
```{r}
sum(!complete.cases(impData))
```
New dataset that is equal to the original dataset but with the missing data

Summary of New dataset
```{r}
summary(impData)
```
Summary of Original dataset
```{r}
summary(data)
```

Histogram of the total number of steps taken each day after imputing the missing vales
```{r}
plot3 <- ggplot(impData, aes(date, steps)) + geom_bar(stat = "identity",binwidth = .5) +
        labs(title = "Histogram of Total Number of Steps Taken Each Day with Imputed Data",x = "Date", y = "Total Number of Steps")
print(plot3)
```

Calculated Mean total number of steps taken per day with Imputed dataset
```{r}
total.steps.impute <- tapply(impData$steps, impData$date, FUN = sum)
mean(total.steps.impute)
```

Calculated Median total number of steps taken per day with Imputed dataset.

```{r}
median(total.steps.impute)

```
Observartions

The mean total number of steps taken per day remained same before and after the imputation

The new Median is slightly higher than the old medain

####Activity patterns between weekdays and weekends
Creating factor variables indicating whether the date is a weekday or weekend
```{r}
impData$weekdays <- factor(format(impData$date, "%A"))
levels(impData$weekdays)
```
```{r}
levels(impData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(impData$weekdays)
```

```{r}
table(impData$weekdays)
```

Panel plot containing time series plot of each 5-minute interval and the average number of steps taken, averaged across all days, split into weekdays and weekends

```{r}
new.averages <- aggregate(impData$steps, 
                      list(interval = as.numeric(as.character(impData$interval)), 
                           weekdays = impData$weekdays),
                      FUN = "mean")
names(new.averages)[3] <- "meanOfSteps"
library(lattice)
plot4 <- xyplot(new.averages$meanOfSteps ~ new.averages$interval | new.averages$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
print(plot4)
```